Interogating the linear variable importance of local explanations of black-box models with animated linear projections

### 2021-11-21 _paper.rmd
Application design and recent screen caps.

@@

@@

```{r globalview, echo=F, out.width = "100%", fig.cap = "Global view screen capture; the approximations of the data and SHAP-spaces of Palmer penguins data. Orientation of the basis contributions is illustrated on a unit circle. Linked brushing allows observations brushed in one plot to be selected in others. This selection is also used in the proceeding view and their corresponding information is displayed in an interactive table. Hovering the cursor over an observation displays a tooltip with row number/name information. In the classification case, misclassified points are circled in red. This view provides an orientation to select a primary and comparison observation, key targets in the following tour."}
knitr::include_graphics("./figures/global_view_penguins.PNG")
```

@@

<!-- Dipiction of a frame -->
The SHAP values of the full dataset are shown as horizontal parrallel coordiate plot at the top with the selected observations highlighted. In the classification case, 1D density curves with underlying rug marks are drawn and colored according to their predicted classes. In the regression case, the horizontal position of the points comes from projection through the 1D attribution basis while the vertical position of is fixed to the observations prediction and residual respectively. The current projection basis is depicted on the bottom, where the width of the bar is mapped to variables contribution to the horizontal axis. The starting frame is the normalized SHAP values of the primary observation.

<!-- cheem tour caption -->
(ref:cheemtour) Cheem tour screen capture; on classification random forest of simulated data, a misclassified observation is be compared with a nearby correctly classified observation. The top contains a horizontal display of the SHAP value from all observation with selected primary and comparison observations highlighted. The middle shows density and rug marks of the current frame with selections similarly highlighted. The bottom shows the projection basis, that is, each variables contribution to the horizontal position. The primary, misclassifed observation (dashed line), is plausibly in the middle of the green density, the story that the explanation trying to sell us. Yet, when we play the tour animating on the contribution of bill length (b_l), the bottom dashed line is more regularly in the center the it observed cluster, orange. By varying this contribution we gain information of how sensitive this variable is to the explanation; the variable importance which led to a misclassification.

```{r cheemtour, echo=F, out.width = "100%", fig.cap = "(ref:cheemtour)"}
knitr::include_graphics("./figures/cheem_tour_penguins.PNG")
```

Data visualization tours animate many linear projections over small changes to the basis. The manual tour creates a basis path by varying the contribution of a selected variable, fully into and out of a projection frame. Doing allows an analyst to test an individual variable's sensitivity to the structure identified in the frame. The default variable selected is the one with the largest discrepancy between the primary and comparison observation's attribution. In the following sections we elaborate on the takeaways we draw from applying this approach in classification and regression tasks respectively.

Another way of thinking about this would be: what would the predictions be if the explanation were different. This is similar to the idea of _ceteris paribus_ profiles [@biecek_ceterisparibus_2020]. _Ceteris paribus_ is Latin for “other things held constant” or “all else unchanged”. The profiles visualize what 'what-if' analysis, showing how an observation's prediction would change from a change in one explanatory variable given that other variables are held constant. In contrast to _ceteris paribus_ profiles, touring methods visualize all observations rather than one, While the variables are orthogonal dimensions, the 2D shadow they cast is not independent. That is, when one variable is rotated out of frame, other variables are effectively rotated into frame.

@@

## Backend infrastructure {#sec:infrastructure}

@@

Provided with tidy and transformed data, the function `cheem_ls` is a wrapper function that facilitates the creation of random forest models and calculates the tree SHAP for all observations. It returns a list containing data frames to be consumed by the application, higher-level model performance, and runtime information. The return of this function can be then be saved to an .rds file for consumption in the application.

## Interactive shiny application

>>>>

<!-- What-if & counterfactual analysis -->
The change in the projection basis might feel similar to counterfactual, what-if analysis, such as _Ceteris paribus_[@biecek_ceterisparibus_2020].  Latin for “other things held constant” or “all else unchanged”, is a counterfactual analysis showing how an observation's prediction would change from a change in one explanatory variable given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation a of the factual observed case by varying the basis, the angle of the data object. Another, but related difference is that this geometric approach maintains orthogonal dimensions. That is to say when the contribution of some variables decrease the the contributions of others necessarily increases.

@@

```{r}
##TODO:XXX
```

@@

```{r globalview, echo=F, out.width = "100%", fig.cap = "Global view screen capture; the approximations of the data and SHAP-spaces of Palmer penguins data. Orientation of the basis contributions is illustrated on a unit circle. Linked brushing allows observations brushed in one plot to be selected in others. This selection is also used in the proceeding view and their corresponding information is displayed in an interactive table. Hovering the cursor over an observation displays a tooltip with row number/name information. In the classification case, misclassified points are circled in red. This view provides an orientation to select a primary and comparison observation, key targets in the following tour."}
knitr::include_graphics("./figures/global_view_toy_class.PNG")
```

@@

<!-- Dipiction of a frame -->
The SHAP values of the full dataset are shown as horizontal parrallel coordiate plot at the top with the selected observations highlighted. In the classification case, 1D density curves with underlying rug marks are drawn and colored according to their predicted classes. In the regression case, the horizontal position of the points comes from projection through the 1D attribution basis while the vertical position of is fixed to the observations prediction and residual respectively. The current projection basis is depicted on the bottom, where the width of the bar is mapped to variables contribution to the horizontal axis. The starting frame is the normalized SHAP values of the primary observation. In figure \@ref(fig:cheemtourclass) we illustrate the classification case with simulated data; 3 clusters separated in the first 2 dimensions with 2 dimensions of noise.

<!-- cheem tour classification caption -->
(ref:cheemtourclass) Cheem tour, classification case. A random forest model predicting the class of the of simulated data. A misclassified observation is be compared with a nearby correctly classified observation. The top shows density and rug marks of the current frame with the positions of the primary and comparison observations shown as dashed and dotted lines. The bottom shows two key features; the distribution of attribution space shown as parallel coordinate lines across the variables and the bars show the current contribution of the variable to the projection. The primary, misclassifed observation (dashed line), is plausibly in the middle of the its prediction,  purple density, the story that the explanation trying to sell us. Yet, when we play the tour animating on the contribution of V1, the bottom dashed line is more regularly in the center the it observed cluster, green, By varying this contribution we gain information of how sensitive this variable is to the explanation; the variable importance which led to a misclassification.

```{r cheemtourclass, echo=F, out.width = "100%", fig.cap = "(ref:cheemtourclass)"}
knitr::include_graphics("./figures/cheem_tour_toy_class.PNG")
```

<!-- tour animation -->
Data visualization tours animate many linear projections over small changes to the basis. The manual tour creates a basis path by varying the contribution of a selected variable, fully into and out of a projection frame. Doing allows an analyst to test an individual variable's sensitivity to the structure identified in the frame. The default variable selected is the one with the largest discrepancy between the primary and comparison observation's attribution. In the following sections we elaborate on the takeaways we draw from applying this approach in classification and regression tasks respectively.

<!-- regression case -->
In the regression case we compare an observation with a particularly extreme residual as compared with a neighboring point that is more  accurately predicted. The global tour is primarily the same without the misclassified circles, but the tooltip now shows the observation's residual. We simulate data, 5 variable from the uniform distribution (between 0 and 5) and $y = x1 + x2 + x1*x2 + (x3 + x4 + x5) / 10 + \epsilon$. We fit a random forest and extract the all tree SHAP values. The first thing we notice from the global view is the shape difference between the data- and attribution-spaces, which are primarily circular are triangular respectively. We also see that the random forest fit the data quite well viewing the predictions across the observations. Though linked brushing we find that the attribution space is highly correlated with the $y$ value. The highest values occupy 1 corner of the triangle, the lowest values cluster on the opposite side and the middle values form a ')' shape filling out the triangle in attribution space. In the regression case the tour view is a bit different, this is illustrated in figure \@ref(cheemtourreg). The the current frames basis and distribution are shown to the left. The horizontal positions of the two scatter plots are the resulting 1D projection of the basis while the heights are fixed to the observed y and the residuals.

<!-- tour regression caption -->
(ref:cheemtourreg) Cheem tour, regression case. A random forest model predicting the continuous $y$ of the of simulated data where $y = x1 + x2 + x1*x2 + (x3 + x4 + x5) / 10 + \epsilon$. The current frames basis and the distribution of all attributions is shown on the left. The horizontal positions for the scatterplots are resulting projection through the 1D basis while the vertical heights are fixed to the observed y and residuals respectively.

```{r cheemtourreg, echo=F, out.width = "100%", fig.cap = "(ref:cheemtourreg)"}
knitr::include_graphics("./figures/cheem_tour_toy_reg.PNG")
```

Now that we have covered the classification and regression cases we will discuss the preprocessing, package infrastructure, and interactive features before proceeding to case studies with data from the wild.

@@

## Package infrastructure {#sec:infrastructure}

@@

<!-- Provided with tidy and transformed data, the function `cheem_ls` is a wrapper function that facilitates the creation of random forest models and calculates the tree SHAP for all observations. It returns a list containing data frames to be consumed by the application, higher-level model performance, and runtime information. The return of this function can be then be saved to an .rds file for consumption in the application. -->

## Interactive features

### 2021-11-21 _paper.tex
Application design and recent screen caps.

@@

{\centering \includegraphics[width=1\linewidth]{./figures/global_view_penguins}

@@

The SHAP values of the full dataset are shown as horizontal parrallel coordiate plot at the top with the selected observations highlighted. In the classification case, 1D density curves with underlying rug marks are drawn and colored according to their predicted classes. In the regression case, the horizontal position of the points comes from projection through the 1D attribution basis while the vertical position of is fixed to the observations prediction and residual respectively. The current projection basis is depicted on the bottom, where the width of the bar is mapped to variables contribution to the horizontal axis. The starting frame is the normalized SHAP values of the primary observation.

@@

{\centering \includegraphics[width=1\linewidth]{./figures/cheem_tour_penguins}

@@

\caption{Cheem tour screen capture; on classification random forest of simulated data, a misclassified observation is be compared with a nearby correctly classified observation. The top contains a horizontal display of the SHAP value from all observation with selected primary and comparison observations highlighted. The middle shows density and rug marks of the current frame with selections similarly highlighted. The bottom shows the projection basis, that is, each variables contribution to the horizontal position. The primary, misclassifed observation (dashed line), is plausibly in the middle of the green density, the story that the explanation trying to sell us. Yet, when we play the tour animating on the contribution of bill length (b\_l), the bottom dashed line is more regularly in the center the it observed cluster, orange. By varying this contribution we gain information of how sensitive this variable is to the explanation; the variable importance which led to a misclassification.}\label{fig:cheemtour}
\end{figure}

@@

Another way of thinking about this would be: what would the predictions be if the explanation were different. This is similar to the idea of \emph{ceteris paribus} profiles (Biecek 2020). \emph{Ceteris paribus} is Latin for ``other things held constant'' or ``all else unchanged.'' The profiles visualize what `what-if' analysis, showing how an observation's prediction would change from a change in one explanatory variable given that other variables are held constant. In contrast to \emph{ceteris paribus} profiles, touring methods visualize all observations rather than one, While the variables are orthogonal dimensions, the 2D shadow they cast is not independent. That is, when one variable is rotated out of frame, other variables are effectively rotated into frame.

@@

\hypertarget{sec:infrastructure}{%
\subsection{Backend infrastructure}\label{sec:infrastructure}}

@@

\hypertarget{preprocessing-1}{%
\subsection{Preprocessing}\label{preprocessing-1}}

Provided with tidy and transformed data, the function \texttt{cheem\_ls} is a wrapper function that facilitates the creation of random forest models and calculates the tree SHAP for all observations. It returns a list containing data frames to be consumed by the application, higher-level model performance, and runtime information. The return of this function can be then be saved to an .rds file for consumption in the application.

\hypertarget{interactive-shiny-application}{%
\subsection{Interactive shiny application}\label{interactive-shiny-application}}

>>>>

The change in the projection basis might feel similar to counterfactual, what-if analysis, such as \emph{Ceteris paribus}(Biecek 2020). Latin for ``other things held constant'' or ``all else unchanged,'' is a counterfactual analysis showing how an observation's prediction would change from a change in one explanatory variable given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation a of the factual observed case by varying the basis, the angle of the data object. Another, but related difference is that this geometric approach maintains orthogonal dimensions. That is to say when the contribution of some variables decrease the the contributions of others necessarily increases.

@@

{\centering \includegraphics[width=1\linewidth]{./figures/global_view_toy_class}

@@

The SHAP values of the full dataset are shown as horizontal parrallel coordiate plot at the top with the selected observations highlighted. In the classification case, 1D density curves with underlying rug marks are drawn and colored according to their predicted classes. In the regression case, the horizontal position of the points comes from projection through the 1D attribution basis while the vertical position of is fixed to the observations prediction and residual respectively. The current projection basis is depicted on the bottom, where the width of the bar is mapped to variables contribution to the horizontal axis. The starting frame is the normalized SHAP values of the primary observation. In figure \ref{fig:cheemtourclass} we illustrate the classification case with simulated data; 3 clusters separated in the first 2 dimensions with 2 dimensions of noise.

@@

{\centering \includegraphics[width=1\linewidth]{./figures/cheem_tour_toy_class}

@@

\caption{Cheem tour, classification case. A random forest model predicting the class of the of simulated data. A misclassified observation is be compared with a nearby correctly classified observation. The top shows density and rug marks of the current frame with the positions of the primary and comparison observations shown as dashed and dotted lines. The bottom shows two key features; the distribution of attribution space shown as parallel coordinate lines across the variables and the bars show the current contribution of the variable to the projection. The primary, misclassifed observation (dashed line), is plausibly in the middle of the its prediction, purple density, the story that the explanation trying to sell us. Yet, when we play the tour animating on the contribution of V1, the bottom dashed line is more regularly in the center the it observed cluster, green, By varying this contribution we gain information of how sensitive this variable is to the explanation; the variable importance which led to a misclassification.}\label{fig:cheemtourclass}
\end{figure}

@@

In the regression case we compare an observation with a particularly extreme residual as compared with a neighboring point that is more accurately predicted. The global tour is primarily the same without the misclassified circles, but the tooltip now shows the observation's residual. We simulate data, 5 variable from the uniform distribution (between 0 and 5) and \(y = x1 + x2 + x1*x2 + (x3 + x4 + x5) / 10 + \epsilon\). We fit a random forest and extract the all tree SHAP values. The first thing we notice from the global view is the shape difference between the data- and attribution-spaces, which are primarily circular are triangular respectively. We also see that the random forest fit the data quite well viewing the predictions across the observations. Though linked brushing we find that the attribution space is highly correlated with the \(y\) value. The highest values occupy 1 corner of the triangle, the lowest values cluster on the opposite side and the middle values form a `)' shape filling out the triangle in attribution space. In the regression case the tour view is a bit different, this is illustrated in figure \ref{cheemtourreg}. The the current frames basis and distribution are shown to the left. The horizontal positions of the two scatter plots are the resulting 1D projection of the basis while the heights are fixed to the observed y and the residuals.

{\centering \includegraphics[width=1\linewidth]{./figures/cheem_tour_toy_reg}

\caption{Cheem tour, regression case. A random forest model predicting the continuous \(y\) of the of simulated data where \(y = x1 + x2 + x1*x2 + (x3 + x4 + x5) / 10 + \epsilon\). The current frames basis and the distribution of all attributions is shown on the left. The horizontal positions for the scatterplots are resulting projection through the 1D basis while the vertical heights are fixed to the observed y and residuals respectively.}\label{fig:cheemtourreg}
\end{figure}

Now that we have covered the classification and regression cases we will discuss the preprocessing, package infrastructure, and interactive features before proceeding to case studies with data from the wild.

@@

\hypertarget{sec:infrastructure}{%
\subsection{Package infrastructure}\label{sec:infrastructure}}

@@

\hypertarget{interactive-features}{%
\subsection{Interactive features}\label{interactive-features}}
